Convective heat transport in stratified atmospheres at low and high Mach number

### 2017-03-10 tex/nusselt_paper.tex
finishes draft of paper

Solutions were time-evolved until a long time average of the fluxes
showed little
variance with depth. By performing a linear stability analysis, 
we determined that the onset of convection in 2D
occurs at $\text{Ra}_{\text{crit}} = \{11.15, 10.06, 10.97, 10.97\}$ 
for $\epsilon = \{1.0, 0.5, 10^{-4}, 10^{-7}\}$ respectively.  Results
in 3D appear to be consistent with our 2D onset values.
We studied Rayleigh
numbers from values at onset to values $\geq 10^6$Ra$_{\text{crit}}$ 
for $\epsilon = 0.5$, up to
$\sim 10^5$Ra$_{\text{crit}}$ at $\epsilon = 10^{-4}$, 
up to $\sim 10^4$Ra$_{\text{crit}}$ for $\epsilon = 1$,
and up to $\sim 10^3$Ra$_{\text{crit}}$ for $\epsilon = 10^{-7}$.

We find that the Mach number is a strong function of 
$\epsilon$ and a weak function of Ra.  
When Ma $< 1$, Ma $\propto \epsilon^{1/2}$, 
but this relation break down as the mean
Ma approaches 1 and the flows become supersonic (see Fig. \ref{fig:ma_v_eps}a).  
When Ma is small,
Ma $\propto$ Ra$^{1/4}$.  At high epsilon, 
this scaling breaks down once the mean atmospheric Ma reaches 1.
Furthermore, the value of $\epsilon$ sets the size of the
evolved thermodynamic variables, 
$T_1/T_0, \rho_1/\rho_0 \propto \epsilon$.  Thus, fluctuations
are very small for low Ma flows, 
but are order 1 when $\epsilon$ and Ma approach 1.

@@

Low Ma flows ($\epsilon = 10^{-4}$)
display the classic narrow downflow and broad upflow lanes of stratified
convection (Fig. \ref{fig:entropy_snapshots}a).
While it has been suggested that pressure forces 
cause symmetry breaking in up- and downflows
\cite{hurlburt&all1984}, at low $\epsilon$ this 
effect seems to be secondary to flows obeying mass conservation as they traverse
the stratified medium.  
Our choice of fixed-temperature, stress free boundary conditions 
allows the flux at the boundaries to vary, while also allowing
for large-scale ``axisymmetric'' horizontal velocities.
Runs at $\text{Ra }> 10^4$
and $\epsilon = 10^{-4}$ exhibit long-term states of
flux disequilibrium, in which roll solutions such as those pictured in 
Fig. \ref{fig:entropy_snapshots}a
are punctuated by shearing states similar to those previously
reported in two-dimensional RB convection \cite{goluskin&all2014}.  
During both of these states, the system is in flux disequilibrium.
A long time average
over both of these states produce flat flux profiles which
can be sensibly analyzed, but the presence of these time-dependent states
makes it difficult to measure fluid quantities in these cases.

At large Ma ($\epsilon = 0.5$), bulk thermodynamic structures are similar but
shock systems form in the upper atmosphere near downflow lanes 
once Ra is sufficiently large  (Fig. \ref{fig:entropy_snapshots}b,c).
Similar shock phenomena were reported in
both two \cite{cattaneo&all1990} and 
three \cite{malagoli&all1990} dimensional polytropic simulations previously.
As Ra is increased to large values 
(Fig. \ref{fig:entropy_snapshots}c), thermodynamic structures 
no longer span the whole domain but rather break up into 
small eddies which traverse the domain multiple
times before diffusing.

@@

The efficiency of convection is quantified by the Nusselt number (Nu).  
Nu is well-defined in RB convection
as the total flux normalized by the steady-state conductive flux 
\cite{johnston&doering2009, otero&all2002}.
In stratified convection Nu is more difficult to define, and we use
a modified version of a traditional stratified Nusselt number 
\cite{graham1975,hurlburt&all1984},
\begin{equation}
\text{Nu} \equiv \frac{\langle F_{\text{conv,z}} + F_{\text{cond,z}} - F_{\text{A}}\rangle}
{\langle F_{\text{cond,z}} - F_{\text{A}}\rangle} 
= 1 + \frac{\langle F_{\text{conv,z}}\rangle}{\langle F_{\text{cond,z}} - F_{\text{A}} \rangle}
\label{eqn:nusselt}
\end{equation}
where $F_{\text{conv,z}}$ and $F_{\text{cond,z}}$ are the 
z-components of $\bm{F}_{\text{conv}}$ and $\bm{F}_{\text{cond}}$,
respectively and $\langle \rangle$ are volume averages.  
$F_{\text{A}} \equiv -\langle\kappa\rangle \partial_z T_{\text{ad}}$ 
is the conductive flux of the proper corresponding adiabatic atmosphere
with $\partial_z T_{\text{ad}} \equiv - g / c_{P}$ 
for a compressible ideal gas in hydrostatic equilibrium \cite{spiegel&veronis1960}.  Here we specify
$\langle \kappa \rangle = \langle \rho\chi \rangle$, which is nearly
$\kappa_0$ when $\epsilon$ is small but can change appreciably for large
values of $\epsilon$.
In incompressible Boussinesq convection, where $\grad S = 0$ only when 
$\grad T = 0$, this definition reduces to the traditional definition
of the Nusselt number \cite{otero&all2002, johnston&doering2009}.

The variation of Nu with Ra is shown in 
Fig. \ref{fig:re_and_nu_v_ra}a.
At low to moderate Ra, 
Nu $\propto$ Ra$^{1/3}$ regardless of $\epsilon$,
as predicted by classical \RB theory \cite{king&all2012}.
At large Ra, Nu $\propto$ Ra$^{1/5}$ 
at $\epsilon = 0.5$.
The scaling of Nu with Ra 
is unclear at low $\epsilon$ once Ra is sufficiently large
for shearing states to occur.  It would be beneficial to study
low $\epsilon$ flows with fixed-flux boundary conditions or
no-slip boundary conditions that disallow these large scale
structures.

The Reynolds number and Peclet number,
\begin{equation}
\text{Re} = \frac{|\bm{u}| L_z}{\nu};\,\,\,\,\text{Pe} = \text{Pr}\,\text{Re},
\end{equation}
quantify the importance of advection to diffusion in the evolved
convective state.  For Pr = 1, such as in this work, Pe = Re.  
Our choice of $\{\nu,\chi\}\propto \rho_0^{-1}$ drastically changes
the value of Re between the top and bottom of the atmosphere.  We report values of
Re at the midplane ($z=L_z/2$) of the atmosphere in
Fig. \ref{fig:re_and_nu_v_ra}b.
At low Ra, Re $\propto$ Ra$^{3/4}$, but at high 
Ra where the average Ma $\approx 1$ for $\epsilon = 0.5$, 
this scaling gives way to a Re $\propto$ Ra$^{1/2}$.
For large values of Ra but at low $\epsilon$, there appears to be a scaling
of Re $\propto$ Ra$^{2/3}$.

@@

As the thermodynamic variables converge to their steady state values, 
the density profile evolves while remaining in hydrostatic equilibrium 
to zeroth order.  In Fig. \ref{fig:nrho_v_ra} we show the number of density 
scale heights present in the evolved solution using two measures.  
We find that once the average Ma of the domain
becomes approximately one, large density inversions begin to 
form in the boundary layers, as was reported by \cite{brandenburg&all2005}.  
Regardless, the agreement of Nu across $\epsilon$, particularly at
low Ra, despite the difference in effective stratifications is striking.

\section{Discussion}
\refstepcounter{section}
\label{sec:discussion}
In this letter we have studied fundamental heat transport by 
stratified convection in simplified 2D and 3D polytropic atmospheres.
We argue that these atmospheres are the natural extension
of the RB problem to stratified systems, 
and are an ideal laboratory for understanding the basic 
properties of stratified convection. 
We see little difference between
the properties of our 2D and 3D results, which aligns with expectations
of Boussinesq theory and DNS results at values of Pr $\geq$ 1 \cite{ahlers&all2009}.

At low Ra and Ma, the scaling of Nu with Ra is reminiscent of RB convection.  
However, at high Ra and Ma, the scaling of Nu becomes weakear, decreasing
from a 1/3 to a 1/5 power law, a decrease by roughly a factor of 1/3.
Similarly, the Reynolds number switches from a 3/4 power law to
a 1/2 scaling in this regime shift, a similar decrease by a factor of 1/3.
This seems to suggest that a large portion of the scaling of the Nusselt number
comes from increases in velocity as Ra increases.  Once the velocity reaches
its natural maximum, heat transport can no longer benefit from increased velocities.
The same is true of the Reynolds number, which tracks increases in velocities as
the flows become more turbulent, but which can only scale with the diffusivities
once the Mach number reaches its maximum value.

Time-dependent oscillating shear states have developed spontaneously, 
as seen before in RB convection \cite{goluskin&all2014}, and these should
be studied in more detail.  These states are fundamentally in flux disequilibrium
and make it very difficult to calculate a meaningful Nusselt number.  It would
be beneficial for future studies to study solutions with fixed flux or no slip
boundary conditions to better converge and study cases at high $\epsilon$ and low
Ra.

Finally, we have found that the stratification of 
these atmospheres evolves in a complex
manner.  Future work should aim to 
understand the importance of stratification on
convective heat transport, and the manner in which
boundary layer density inversions interact with
thermal plumes.

>>>>

Solutions were time-evolved until a long time average of the fluxes
showed little
variance with depth. By performing a linear stability analysis, 
we determined that the onset of convection in 2D
occurs at $\text{Ra}_{\text{crit}} = \{11.15, 10.06, 10.97, 10.97\}$ 
for $\epsilon = \{1.0, 0.5, 10^{-4}, 10^{-7}\}$, respectively.  These
onset values hold in 3D.
We studied Rayleigh
numbers $\lesssim \left\{10^7, 10^5, 10^4, 10^3\right\}$Ra$_{\text{crit}}$ 
for $\epsilon = \{0.5, 10^{-4}, 1.0, 10^{-7}\}$, respectively.

We find that the Mach number is a strong function of 
$\epsilon$ and a weak function of Ra.  
When Ma $< 1$, we observe a scaling law of
Ma(Ra$, \epsilon) \propto \epsilon^{1/2}$Ra$^{1/4}$, 
but this relation breaks down as the mean
Ma approaches 1 and the flows become supersonic (see Fig. \ref{fig:ma_v_eps}).  
Furthermore, the value of $\epsilon$ sets the size of the
evolved thermodynamic variables, such that
$T_1/T_0 \propto \epsilon$ and $\rho_1/\rho_0 \propto \epsilon$.  
These fluctuations are very small when $\epsilon$ and Ma are small,
but can be O(1) for values of $\epsilon$ near $m_{\text{ad}}$.

@@

Low Ma flows (e.g. $\epsilon = 10^{-4}$)
display the classic narrow downflow and broad upflow lanes of stratified
convection (Fig. \ref{fig:entropy_snapshots}a).
While it has been suggested that buoyancy breaking as 
a result of pressure forces leads to
the asymmetric nature of up- and downflows
\cite{hurlburt&all1984}, at low $\epsilon$ this 
effect seems to be unimportant, and the convective structures form
as a result of flows obeying mass conservation while traversing
the stratified medium.  
Our choice of fixed-temperature, stress free boundary conditions 
allows the flux at the boundaries to vary, while also allowing
for domain-wide horizontal velocity profiles along the boundaries.
2D runs at $\text{Ra }> 10^3$Ra$_{\text{crit}}$
and $\epsilon = 10^{-4}$ exhibit long-term states of
flux disequilibrium, in which roll solutions such as those pictured in 
Fig. \ref{fig:entropy_snapshots}a
are punctuated by shearing states similar to those previously
reported in two-dimensional RB convection \cite{goluskin&all2014}.  
During both of these states, the system is in flux disequilibrium.
A long time average
over both of these states produces a flat flux profile which
can be sensibly analyzed, but the presence of these time-dependent states
makes it difficult to measure fluid quantities in these higher-Ra
cases at low $\epsilon$.

At high Ma (e.g. $\epsilon = 0.5$), bulk thermodynamic structures are similar but
shock systems form in the upper atmosphere near downflow lanes 
for sufficiently large values of $\epsilon$ and Ra (Fig. \ref{fig:entropy_snapshots}b,c).
Similar shock phenomena were reported in
both two \cite{cattaneo&all1990} and 
three \cite{malagoli&all1990} dimensional polytropic simulations previously.
As Ra is increased to large values 
(Fig. \ref{fig:entropy_snapshots}c), thermodynamic structures 
no longer span the whole domain but rather break up into 
small eddies which traverse the domain multiple
times before diffusing.

@@

The efficiency of convection is quantified by the Nusselt number (Nu).  
Nu is well-defined in RB convection
as the total flux normalized by the steady-state conductive flux 
\cite{johnston&doering2009, otero&all2002}.
In stratified convection Nu is more difficult to define, and we use
a modified version of a traditional stratified Nusselt number 
\cite{graham1975,hurlburt&all1984},
\begin{equation}
\text{Nu} \equiv \frac{\langle F_{\text{conv,z}} + F_{\text{cond,z}} - F_{\text{A}}\rangle}
{\langle F_{\text{cond,z}} - F_{\text{A}}\rangle} 
= 1 + \frac{\langle F_{\text{conv,z}}\rangle}{\langle F_{\text{cond,z}} - F_{\text{A}} \rangle}
\label{eqn:nusselt}
\end{equation}
where $F_{\text{conv,z}}$ and $F_{\text{cond,z}}$ are the 
z-components of $\bm{F}_{\text{conv}}$ and $\bm{F}_{\text{cond}}$,
and $\langle \rangle$ are volume averages.  
$F_{\text{A}} \equiv -\langle\kappa\rangle \partial_z T_{\text{ad}}$ 
is the conductive flux of the proper corresponding adiabatic atmosphere in
in thermal equilibrium.
For a compressible ideal gas in hydrostatic equilibrium,
$\partial_z T_{\text{ad}} \equiv - g / c_{P}$ \cite{spiegel&veronis1960}.  
It is important to measure the evolved value of
$\langle \kappa \rangle = \langle \rho\chi \rangle$, which is nearly
$\kappa_0$ when $\epsilon$ is small but can change appreciably for large
values of $\epsilon$.
In incompressible Boussinesq convection, where $\grad S = 0$ only when 
$\grad T = 0$, this definition reduces to the traditional definition
of the Nusselt number \cite{otero&all2002, johnston&doering2009}.

The variation of Nu with Ra is shown in 
Fig. \ref{fig:re_and_nu_v_ra}a.
At low to moderate Ra, 
Nu $\propto$ Ra$^{1/3}$ regardless of $\epsilon$,
reminiscent of scaling laws in classical \RB theory \cite{king&all2012}.
At large Ra, Nu $\propto$ Ra$^{1/5}$ 
at $\epsilon = 0.5$.
The scaling of Nu with Ra 
is unclear at low $\epsilon$ once Ra is sufficiently large
for shearing states to occur.  It would be beneficial to study
low $\epsilon$ flows with fixed-flux boundary conditions or
no-slip boundary conditions that disallow time-dependent shear states.

The Reynolds number (Re = $|\bm{u}|L_z/\nu$) and Peclet number
(Pe = Pr Re)
compare the importance of advection to diffusion in the evolved
convective state.  For Pr = 1, such as in this work, Pe = Re.  
Our choice of $\{\nu,\chi\}\propto \rho_0^{-1}$ drastically changes
the value of Re between the top and bottom of the atmosphere.  We report values of
Re at the midplane ($z=L_z/2$) of the atmosphere in
Fig. \ref{fig:re_and_nu_v_ra}b.
At low Ra, Re $\propto$ Ra$^{3/4}$.  At high Ra and $\epsilon$, 
where the average Ma $\approx 1$, 
this scaling gives way to a Re $\propto$ Ra$^{1/2}$,
where all changes in Re are due to the lowering of the diffusivity
through raising Ra.
For large values of Ra but at low $\epsilon$, there appears to be a scaling
of Re $\propto$ Ra$^{2/3}$, but once again the shearing states have made
these points difficult to measure.

@@

As the thermodynamic variables converge to their steady state values, 
the density profile evolves while remaining in hydrostatic equilibrium 
to zeroth order.  In Fig. \ref{fig:nrho_v_ra} we show the number of density 
scale heights present in the evolved solution using two measures.  
We find that once the average Ma of the domain
becomes approximately one, large density inversions begin to 
form in the boundary layers, as was reported by \cite{brandenburg&all2005}.  
The agreement of Nu across $\epsilon$ (Fig. \ref{fig:re_and_nu_v_ra}a), 
particularly at low Ra in which all four of our cases collapse onto a single
power law, is striking in light of the vastly different evolved stratifications
felt by the flows.

\section{Discussion}
\refstepcounter{section}
\label{sec:discussion}
In this letter we have studied fundamental heat transport by 
stratified convection in simplified 2D and 3D polytropic atmospheres.
We argue that these atmospheres are the natural extension
of the RB problem to stratified systems, 
and are an ideal laboratory for understanding the basic 
properties of stratified convection. 
We see little difference between
the properties of our 2D and 3D results, which aligns with expectations
of Boussinesq theory and numerical simulation results at 
values of Pr $\geq$ 1 \cite{ahlers&all2009}.

At low Ra and Ma, the scaling of Nu with Ra is reminiscent of RB convection.  
However, at high Ra and Ma, the scaling of Nu becomes weaker, changing
from a 1/3 to a 1/5 power law, a decrease by nearly a factor of 1/3.
Similarly, the Reynolds number switches from a 3/4 power law to
a 1/2 scaling in this regime shift, a similar decrease.
This seems to suggest that a large portion of the scaling of the Nusselt number
and Reynolds number in the low-Ma regime
comes from increases in velocity.  Once the velocity reaches
its natural maximum, heat transport can no longer benefit from increased velocities.
The same is true of the Reynolds number, which scales directly with system diffusivities
once the Mach number reaches its maximum value.

Time-dependent oscillating shear states have developed spontaneously
at values of $\epsilon=10^{-4}$ and Ra $\gtrsim 10^3$Ra$_{\text{crit}}$ in our 2D simulations.  
Similar states have been observed in 2D RB convection \cite{goluskin&all2014}, and these should
be studied in more detail.  These states are fundamentally in flux disequilibrium
and last for large time scales (O(100$t_b$)),
making it very difficult to choose a proper time window over which to calculate a 
meaningful Nusselt number.  It would
be beneficial for future studies to examine atmosperes with fixed flux or no slip
boundary conditions to better converge and study cases at low $\epsilon$ and high
Ra.

Finally, we have found that the stratification of 
these atmospheres evolves in a complex
manner.  Future work should aim to 
understand the importance of stratification on
convective heat transport, and the manner in which
boundary layer density inversions interact with
the formation and transport properties of thermal plumes.
